---
title: "Capstone Project"
author: "Roderick Fan"
date: "2023-03-29"
output: html_document
---

Clean Environment
Loading Packages, dataset
```{r}
rm(list = ls())
cat("\14")
library(tidyverse)
library(here)
library(party)
library(lubridate)
library(lme4)
library(randomForest)
library(caret)
library(rpart)
library(rpart.plot)
library(car)
library(e1071)
library(sqldf)
library(ROSE)
spot <- read.csv(here("Spot Data.csv"))
```



Preprocessing
```{r}
#Rename Patient ID
spot$Patient.ID <- as.numeric(factor(spot$Patient.ID, levels=unique(spot$Patient.ID)))

#Binomial Sex
spot$Patient.Sex <- as.factor(spot$Patient.Sex)


#Dropping Redundant Predictors   
spot <- spot[,-c(5)]

#Factoring Specimen Type
spot$Specimen.Type <- as.factor(spot$Specimen.Type)

#Factoring and Renaming Client Location
spot$Client.ID <- as.factor(spot$Client.ID)
levels(spot$Client.ID) <- c("Location A","Location B","Location C","Location D","Location E","Location F",
                            "Location G","Location H","Location I","Location J","Location K","Location L",
                            "Location M","Location N","Location O","Location P","Location Q","Location R")

table(spot$Client.ID)
#Convert and Simplify Datetime Columns
datetime_cols <- c("Datetime.Kit.Delivered", "Datetime.Kit.Registered", "Datetime.Sample.Sent.In", "Datetime.Sample.Received", "Datetime.of.End.Status")
for (col in datetime_cols) {
  spot[[col]] <- as.Date(substr(spot[[col]], 1, 10), format="%Y-%m-%d")
}
```


```{r}

#Calculate Time Gap between Sent in and Resulted
spot$time.gap1 <- as.numeric(difftime(spot$Datetime.Sample.Sent.In, spot$Datetime.Kit.Delivered,units = "day"))
spot$time.gap2 <- as.numeric(difftime(spot$Datetime.Sample.Received, spot$Datetime.Sample.Sent.In,units = "day"))
spot$time.gap3 <- as.numeric(difftime(spot$Datetime.of.End.Status, spot$Datetime.Sample.Received,units = "day"))

#spot.clean <- spot[,-c(7:11)]
#spot.clean <- na.omit(spot.clean)
spot$End.Status[spot$End.Status == "partially_resulted"] <- "resulted"
spot$End.Status[spot$End.Status %in% c("collected", "collection_exception","delivered","delivery_exception","received")] <- "rejected"

test.rate <- spot%>%
  group_by(Patient.ID,Specimen.Type) %>%
  summarise(Count = n(),
            resulted_rate = sum(End.Status == "resulted")/n(),
            gap1 = mean(time.gap1,na.rm = T),
            gap2 = mean(time.gap2,na.rm = T),
            gap3 = mean(time.gap3,na.rm = T))


```


Customer Age Distribution
```{r}
ggplot(spot, aes(x =spot$Patient.Age, fill = Patient.Sex)) +
  geom_histogram(color = "black", alpha = 0.5, binwidth = 5) +
  scale_fill_manual(values = c("#1F78B4","pink"),
                   labels = c("Male", "Female"))+
  theme(panel.background = element_rect(fill = "white"),legend.position = "right")+
  scale_x_continuous(breaks = seq(15, 90, by = 5))+
  labs(title = "Age Distribution of Patients",
       x = "Patient Age",
       y = "Number of Patients",
       fill = "")
```


Geographical Information of Kits Delivery
```{r}
# Tableau
```


calculate time gap and visualize the distribution of time gap
```{r}
spot.clean <- spot[,-c(7:11)]
spot.clean <- na.omit(spot.clean)
spot.clean$End.Status[spot.clean$End.Status == "partially_resulted"] <- "resulted"
spot.clean$End.Status[spot.clean$End.Status %in% c("collected", "collection_exception","delivered","delivery_exception","received")] <- "rejected"

threshold.1 <-  1.5*(IQR(spot.clean$time.gap1))
outliers.1 <- spot.clean$time.gap1 < median(spot.clean$time.gap1) - threshold.1 | spot.clean$time.gap1 > median(spot.clean$time.gap1) + threshold.1
spot.gap.1 <- spot.clean[!outliers.1,]

threshold.3 <-  1.5*(IQR(spot.clean$time.gap3))
outliers.3 <- spot.clean$time.gap3 < median(spot.clean$time.gap3) - threshold.3 | spot.clean$time.gap3 > median(spot.clean$time.gap3) + threshold.3
spot.gap.3 <- spot.clean[!outliers.3,]

#spot.gap.3$Specimen.Type <- factor(spot.gap.3$Specimen.Type, labels = c("stool","saliva_3", "adx100", "adx100_4", "dry_tranport_swap",  "whatman903_10","adx100_2"))


box.gap.1 <- ggplot(data = spot.gap.1)+
                geom_boxplot(aes(x = Patient.Sex, y = time.gap1, fill = Patient.Sex), alpha = 0.5)+
                scale_fill_manual(values = c("pink","#1F78B4"),
                                  labels = c("Female", "Male"))+
                theme(panel.background = element_rect(fill = "white"),legend.position = "top")+
                coord_flip()+
                labs(x = "Patient Sex",
                     y = "Time Between Delivered to Sent in",
                     fill = "",
                     title = "Time Gap Comparison")


box.gap.3 <- ggplot(data = spot.gap.3)+
                geom_boxplot(aes(x = Specimen.Type, y = time.gap3, fill = Specimen.Type), alpha = 0.3)+
                theme(panel.background = element_rect(fill = "white"),legend.position = "none")+
                coord_flip()+
                labs(x = "Specimen Types",
                     y = "Time Between Received by Test Center and Resulted",
                     fill = "",
                     title = "Time Gap Comparison")




box.gap.1
box.gap.3


```



visualize the Distribution of Number of Tests Taken by Patients
```{r}
test.duration <- spot%>%
  count(Patient.ID) 

test.duration$n <- as.factor(test.duration$n)
test.duration <- test.duration[!is.na(test.duration$n),]

ggplot(test.duration, aes(x = n)) +
  geom_bar(alpha = 0.5, color = "black", fill = "lightblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), hjust = -0.5) +
  ylim(0,3500)+
  coord_flip()+
  labs(x = "Number of Tests Taken", 
       y = "Count",
       title ="Distribution of Number of Tests Taken by Patients") +
  theme_classic()

```


Visualization of Success Rate and Count by Specimen Type
```{r}
test.rate <- spot.clean%>%
  group_by(Specimen.Type) %>%
  summarize(Count = n(), resulted_rate = sum(End.Status == "resulted")/n())

ggplot(test.rate, aes(x = reorder(Specimen.Type,Count), y = resulted_rate, fill = Specimen.Type)) +
  geom_bar(aes(y = Count),stat = "identity", alpha = 0.5, width = 0.5) +
  geom_text(aes(label = paste0(Count, " / ",round(resulted_rate*100,2), "%"), y = Count + 1), 
            color = "black", fontface = "bold", size = 4, hjust = -0.1) +
  labs(x = "", y = "Test times (with Success Rate)", 
       title = "Success Rate and Count by Specimen Type") +
  scale_fill_discrete(name = "Specimen Type") +
  scale_y_continuous(sec.axis = sec_axis(~./max(test.rate$count), name = "Count"))+
  ylim(0,8000)+
  coord_flip()+
  theme_classic()+
  theme(legend.position = "none")

```


Visualization of Success rate and Count by Client.ID
```{r}

test.rate <- spot.clean%>%
  group_by(Client.ID) %>%
  summarize(Count = n(), resulted_rate = sum(End.Status == "resulted")/n())

ggplot(test.rate, aes(x = reorder(Client.ID,Count), y = resulted_rate, fill = Client.ID)) +
  geom_bar(aes(y = Count),stat = "identity", alpha = 0.5, width = 0.5) +
  geom_text(aes(label = paste0(Count, " / ",round(resulted_rate*100,2), "%"), y = Count + 1), 
            color = "black", fontface = "bold", size = 4, hjust = -0.1) +
  labs(x = "", y = "Test times (with Success Rate)", 
       title = "Success Rate and Count by Client ID") +
  scale_fill_discrete(name = "Client ID") +
  scale_y_continuous(sec.axis = sec_axis(~./max(test.rate$count), name = "Count"))+
  ylim(0,8000)+
  coord_flip()+
  theme_classic()+
  theme(legend.position = "none")


```




Model Build: Random Forest
```{r}
set.seed(111)
t(t(colnames(spot.clean)))

# partition
set.seed(1)
spot.df <- spot.clean[,-c(2,6,8)]
train.index <- sample(c(1:dim(spot.df)[1]), dim(spot.df)[1]*0.6)
train <- spot.df[train.index,]
valid <- spot.df[-train.index,]

# random forest
spot.rf <- randomForest(as.factor(End.Status) ~ .-Patient.ID, data = train, ntree = 200, mtry=4, nosize=5, importance=TRUE)


x <- ctree(as.factor(End.Status) ~ Patient.Sex + Patient.Age + Specimen.Type + time.gap1 + time.gap2 + time.gap3, data = spot.df)

plot(x)

varImpPlot(spot.rf, type=1, main = "Random Forest Variable Importance Plot")
spot.rf.pred <- predict(spot.rf, valid)
confusionMatrix(spot.rf.pred, as.factor(valid$End.Status), positive = "resulted")

```


Naive Bayes
```{r}
#Set Naive Bayes Data set
naive.df <- spot.clean
naive.df$Patient.Sex <- ifelse(naive.df$Patient.Sex == "M",1,0)
naive.df$End.Status <- ifelse(naive.df$End.Status == "resulted", 1, 0)

naive.index <- sample(c(1:dim(naive.df)[1]), dim(naive.df)[1]*0.6)
naive.train <- naive.df[naive.index,]
naive.valid <- naive.df[-naive.index,]


model1 <- naiveBayes(End.Status ~ ., data = naive.train)
model2 <- naiveBayes(End.Status ~ . - Patient.ID - time.gap2, data = naive.train)
model3 <- naiveBayes(End.Status ~ . - Patient.ID - time.gap2 - Patient.Zip, data = naive.train)
model4 <- naiveBayes(End.Status ~ . - Patient.ID - time.gap2 - Patient.Zip - Rejection.Reason, data = naive.train)

predictions1 <- predict(model1, naive.valid)
predictions2 <- predict(model2, naive.valid)
predictions3 <- predict(model3, naive.valid)
predictions4 <- predict(model4, naive.valid)

CM1 <- confusionMatrix(table(predictions1, naive.valid$End.Status), positive = "1")
CM2 <- confusionMatrix(table(predictions2, naive.valid$End.Status), positive = "1")
CM3 <- confusionMatrix(table(predictions3, naive.valid$End.Status), positive = "1")
CM4 <- confusionMatrix(table(predictions4, naive.valid$End.Status), positive = "1")

CM1
CM2
CM3
CM4
```


```{r}
#Rejection Reason is a key indicator

library(pROC)

roc4 <- roc(predictions4, naive.valid$End.Status, levels = c(0,1))


auc4 <- round(auc(roc4), 3)

cat("AUC for Model 4: ", auc4, "\n")



```



Logistic Regression Model
```{r}

table(spot.mix$End.Status)

spot.mix <- spot.clean[,c(1,3,4,5,7,8,9,10,11)]
spot.mix$End.Status <- ifelse(spot.mix$End.Status == "resulted", 1, 0)
spot.mix$Rejection.Reason[spot.km$Rejection.Reason == ""] <- "No Reason"
spot.mix$Rejection.Reason[spot.km$Rejection.Reason == "qns\\"] <- "qns"
spot.mix$Rejection.Reason <- as.factor(spot.mix$Rejection.Reason)

spot.mix <- ROSE(as.factor(End.Status) ~ ., data = spot.mix)$data

table(spot.mix$End.Status)
train.index.m <- sample(c(1:dim(spot.df)[1]), dim(spot.df)[1]*0.6)
train.mix <- spot.mix[train.index.m,]
valid.mix <- spot.mix[-train.index.m,]

mixed.model <- glmer(End.Status ~  (1|Specimen.Type) + (1|Patient.Sex)
                                  +  time.gap1
                                  +  time.gap2
                                  +  time.gap3
                                  +  Patient.Age,
                     data = train.mix,
                     family = binomial)

summary(mixed.model)
test_predictions <- predict(mixed.model,valid.mix,type = "response")

confusionMatrix(as.factor(ifelse(test_predictions>0.5,1,0)),as.factor(valid.mix$End.Status), positive = "1")

roc3 <- roc(valid.mix$End.Status,test_predictions)

spot.rf.pred <- as.ordered(spot.rf.pred)
roc.rf <- roc(valid$End.Status,spot.rf.pred)


plot(roc3, col = "tomato", main = "ROC Curves Comparison")
lines(roc4, col = "cornflowerblue")
lines(roc.rf, col = "gold")

legend("bottomright", c("Multi-Level Regression","Random Forest","Naive Bayes"), col = c("tomato", "gold", "cornflowerblue"), lty = 1)


auc.m <- round(auc(roc3), 3)
auc.n <- round(auc(roc4), 3)
auc.r <- round(auc(roc.rf), 3)
cat("AUC for Multi-Level Regression: ", auc.m, "\n")
cat("AUC for Random Forest: ", auc.r, "\n")
cat("AUC for Naive Bayes: ", auc.n, "\n")
```


```{r}
logit1 <- glm(as.factor(End.Status) ~ time.gap2+time.gap3, data = spot.df, family = "binomial")
summary(logit1)
```

```{r}
spot.km <- subset(spot, End.Status == "rejected")
spot.km$Rejection.Reason[spot.km$Rejection.Reason == ""] <- "No Reason"
spot.km$Rejection.Reason[spot.km$Rejection.Reason == "qns\\"] <- "qns"


spot.km <- spot.km[,-c(7:11)]
#spot.km$time.gap1[is.na(spot.km$time.gap1)] <- mean(spot.km$time.gap1, na.rm = TRUE)
#spot.km$time.gap2[is.na(spot.km$time.gap2)] <- mean(spot.km$time.gap2, na.rm = TRUE)




spot.km.1 <- spot.km%>%
  subset(spot.km$Rejection.Reason != "No Reason")

table(spot.km.1$Rejection.Reason)

ggplot(spot.km.1, aes(x =spot.km.1$Patient.Age, fill = Patient.Sex)) +
  geom_density(alpha = 0.2) +
  scale_fill_manual(values = c("#1F78B4","pink"),
                   labels = c("Male", "Female"))+
  theme(panel.background = element_rect(fill = "white"),legend.position = "right")+
  scale_x_continuous(breaks = seq(15, 90, by = 10))+
  facet_wrap(~Rejection.Reason)+
  labs(title = "Proportion of Patients",
       x = "Patient Age",
       y = "Number of Patients",
       fill = "")


table(spot.km.1$Specimen.Type)
spot.km.2 <- spot.km.1 %>% 
                  group_by(Rejection.Reason,Specimen.Type) %>%
                      summarise(count = n()) %>% 
                      mutate(ratio = count / sum(count))

#spot.km.2$Specimen.Type <- factor(spot.km.2$Specimen.Type, labels = c("adx100_2","adx100_4","adx100"))

ggplot(data = spot.km.2, aes(x = Rejection.Reason, y = ratio, fill=Rejection.Reason)) +
  geom_bar(stat = "identity", color = 'black')+
  geom_text(aes(label = paste0(round(ratio*100), "%"), y = ratio),vjust = -1)+
  ylim(0,1)+
  facet_wrap(~Specimen.Type)+
  theme(panel.background = element_rect(fill = "white"),legend.position = "none")+
  scale_fill_brewer(palette="Paired")
```







